---
title: "DPMs for Non-Monotone Missingness"
output: html_document
editor_options: 
  markdown: 
    wrap: 80
---

# Introduction

In this document we analyze data synthetic data to illustrate the application of
the Dirichlet process mixture of multinomial models to estimate the mean of an
outcome $Y_{iJ}$ at the dnof a study, assuming that the outcome is measured at
times $j = 1,\ldots, J$ with (possibly) non-monotone missingness. We show both
how to use this method for fully-Bayesian inference and how to use it as part of
a multiple imputation (MI) approach.

For simplicity we consider only a single treatment arm. To compare two
treatments, one can fit separate models to each treatment and then combine the
results from the different imputations and/or samples from the posterior to
perform inference for the differences in means.

# Description of the Observed Data Model

The Dirichlet process mixture of multinomial models for the observed data can be
understood as a mixture model of the form

$$
  f(y_r, r)
    = \sum_k w_k 
      \left\{\prod_{j=1}^J \gamma_{kj}^{r_j} (1 - \gamma_{kj})^{1 - r_j}\right\}
      \left\{\prod_{j : r_j = 1} \beta_{kj}^{y_j} 
          (1 - \beta_{kj})^{1 - y_j}\right\}.
$$

This model can be described in a hierarchical fashion as follows:

$$
\begin{aligned}
  Z_i &\sim \text{Categorical}(w), \\
  [R_{ij} \mid Z_i = k] &\sim \text{Bernoulli}(\gamma_{kj}), \\
  [Y_{ij} \mid R_{ij} = 1, Z_i = k] &\sim \text{Bernoulli}(\beta_{kj}). \\
\end{aligned}
$$

The parameters
$(w_k, \gamma_{kj}, \beta_{kj} : k = 1,\ldots,K, j = 1, \ldots, J)$ are then
specified according to a *truncated Dirichlet process prior* with

$$
\begin{aligned}
  w_k &= w'_k \prod_{\ell < k} (1 - w'_\ell), 
   \qquad w'_k \sim \text{Beta}(1,\alpha), \\
  \beta_{kj} &\sim \text{Beta}\{\rho_{\beta_j} \lambda_{\beta_j}, 
                               (1 - \rho_{\beta_j})\lambda_{\beta_j}\}, \\
  \gamma_{kj} &\sim \text{Beta}\{\rho_{\gamma_j} \lambda_{\gamma_j}, 
                               (1 - \rho_{\gamma_j})\lambda_{\gamma_j}\}. \\
\end{aligned}
$$

This approach essentially sets all outcomes to be independent of one another
conditional on a single latent class variable $Z_i$; the $\rho$ parameters allow
for the overall (marginal) probabilities of the variables to be similar across
latent classes, while the $\lambda$ parameters are precision parameters that
allow the probabilities to vary across clusters. These parameters are given
independent hyperpriors of the generic form

$$
\begin{aligned}
  \rho &\sim \text{Uniform}(0,1), \\
  \pi(\lambda) &\sim \frac{\sigma}{(\sigma + \lambda)^2}
\end{aligned}
$$

where $\sigma$ is taken to be a "prior guess" at the number of observations in a
typical latent class. As shown in the book, this corresponds to a type of
*uniform shrinkage prior* for the estimated mean within each cluster. A very
rough guess at this quantity, assuming a fixed $\alpha$, is
$\frac{N}{\alpha\{\psi(N + \alpha) - \psi(\alpha)\}}$ where $\psi(x)$ is the
digamma function, although it also makes sense to try different values of
$\sigma$ to assess how sensitive the results are to its choice (generally, we
have not found it to be highly influential). Alternatively, one might consider
placing a prior on $\sigma$, although the current software does not allow for
this.

# Data Generation

The data used in Chapter 12 is proprietary; consequently, we illustrate the use
of the software using synthetic data. This is generated below:

```{r setup, results='hide'}
## Load ----
library(tidyverse)
library(CausalBNPBook)
library(bindata)
library(mitools)
library(tidybayes)

## Generate data ---

set.seed(1234)
simulate_data <- function(N = 500, J = 6, rho = .4) {
  Sigma <- (1 - rho) * diag(J) + rho
  Y <- rmvbin(N, margprob = rep(.5, J), sigma = Sigma)
  R <- rmvbin(N, margprob = rep(.8, J), sigma = Sigma)
  Y <- ifelse(R == 1, Y, NA)

  return(list(Y = Y, R = R))
}

synthetic_data <- simulate_data()
```

In this case, $Y_i$ and $R_i$ are generated independently (hence, missingness is
missing completely at random) according to a multivariate probit model with an
exchangeable covariance $\Sigma_{ij} = 0.4 + 0.6 I(i = j)$. For more details on
the generation of the data, see `?rmvbin`.

# Fitting the Model to the Observed Data

The main function for fitting the Dirichlet mixture of multinomials model is
`ParafacMNAR`, which returns an object of class `parafac`. A missing-at-random
version of the model can be fit using the `ParafacMAR` model; this uses the same
data generating mechanism for the observed data for $Y_{ij}$ but does not
explicitly model the missing data indicators. The arguments for fitting the
model are self explanatory, with `sigma_a` corresponding to the parameter
$\sigma$ in the uniform shrinkage prior.

```{r fitobs, cache = TRUE, results='hide'}
set.seed(7483)

fit_mnar <- ParafacMNAR(
  Y       = synthetic_data$Y,
  R       = synthetic_data$R,
  sigma_a = 8,
  nburn   = 1000,
  nsave   = 1000,
  nthin   = 1
)

fit_mar <- ParafacMAR(
  Y       = synthetic_data$Y,
  R       = synthetic_data$R,
  sigma_a = 8,
  nburn   = 1000,
  nsave   = 1000,
  nthin   = 1
)
```

We will not generally interact with the output of these functions directly;
instead, they are used in conjunction with either multiple imputation or with
the $g$-computation functions to sample from the posterior of the parameters
$\psi_j = \mathbb{E}(Y_{ij})$.

# Performing Multiple Imputation

As in the monograph, this software allows for specification of several different
identifying assumptions. The function `ParafacMI` allows us to draw multiply
imputed datasets under a variety of different assumptions, including the
complete case missing value (CCMV), nearest identified pattern (NIP), and tilted
last occasion (TLO) assumptions discussed in the book. The tilted last occasion
model allows for specifying a prior on a sensitivity parameter $\xi_j$, which is
discussed below for performing a sensitivity analysis; currently, TLO is the
only assumption that allows specifying a sensitivity parameter. The code below
performs the imputations under CCMV, NIP, and MAR:

```{r impute, cache = TRUE}
set.seed(7371)

imputes_ccmv <- ParafacMI(Y = synthetic_data$Y,
                          R = synthetic_data$R,
                          chain = fit_mnar,
                          num_impute = 20,
                          method = "CCMV")

imputes_nip <- ParafacMI(Y = synthetic_data$Y,
                         R = synthetic_data$R,
                         chain = fit_mnar,
                         num_impute = 20,
                         method = "NIP",
                         j_0 = 6)

imputes_mar <- ParafacMI(Y = synthetic_data$Y,
                         R = synthetic_data$R,
                         chain = fit_mar,
                         num_impute = 20,
                         method= "MAR")
```

This returns a list of imputed outcomes `Y` in a "wide" format, which we now
inspect:

```{r}
head(imputes_ccmv[[1]])
head(imputes_nip[[1]])
```

Note that the imputations under the nearest identified pattern (NIP) assumption
still contains `NA`'s in the imputation; this is because NIP is a *partial*
identifying restriction as defined in the monograph, with only a single time
point being imputed to ensure that the analyst does not attempt to draw any
conclusions that are not supported by the assumption.

These imputed datasets can be used with existing tools for multiple imputation
in the usual way. For example, using the `mitools` package, we can perform
inference for the mean outcome at time $J$ under the CCMV assumption as follows
(see `?MIcombine` for details on how to combine inferences across different
imputations):

```{r mi, cache = TRUE}
## Collect a list containing point estimates of the mean outcome at time J
mean_ccmv <- lapply(imputes_ccmv, function(x) mean(x[,ncol(x)]))

## Collect a list containing standard errors of the point estimate
var_ccmv <- lapply(imputes_ccmv, function(x) var(x[,ncol(x)]) / nrow(x))

## Combine the results via multiple imputation
MIcombine(mean_ccmv, var_ccmv)
```

# Fully-Bayes Inference

The `GcompCCMV` and `GcompODMV` functions implement the $g$-formula to produce
samples of the mean at time $j$ under the CCMV and NIP assumptions,
respectively. Both of these functions allow for specifying a prior for a
sensitivity parameter $\xi_j$, which is analogous to the exponential tilting
parameter used in tilted last occasion model described in the monograph. These
functions also return an estimate of the Monte Carlo error for each iteration,
so that the user can judge whether the Monte Carlo error from the
$g$-computation is sufficiently small to treat the samples as exact samples from
the posterior or use these standard errors with the accelerated $g$-computation
algorithm.

```{r gcomp, cache = TRUE}
set.seed(99991)

mu_ccmv <- GcompCCMV(fit_mnar)
mu_nip <- GcompODMV(fit_mnar, j_0 = 6)
```

These objects are returned in a tidy format, with the iteration, sample, and
standard error as the columns:

```{r}
head(mu_ccmv)
```

The results can be summarized, for example, using the `tidybayes` library to
visualize the posterior distribution and credible intervals for the parameters.
The following code compares the inferences for the two assumptions:

```{r}
mu_ccmv_2 <- mu_ccmv %>% mutate(Assumption = "CCMV")
mu_nip_2 <- mu_nip %>% mutate(Assumption = "NIP")

mu_compare <- rbind(mu_ccmv_2, mu_nip_2)

ggplot(mu_compare, aes(x = mean, y = Assumption, color = Assumption)) +
  stat_halfeye(alpha = 0.5) +
  theme_bw() +
  xlab("Mean of Y")
```

and also examines how the samples differ from each other on an
iteration-by-iteration basis:

```{r}
mu_delta  <- mu_ccmv %>% mutate(delta = mean - mu_nip$mean)
ggplot(mu_delta, aes(x = delta)) +
  stat_halfeye(alpha = 0.5, color = 2, size = 5) +
  theme_bw()
```

# Incorporating Sensitivity Parameters

In the current version of this package, there is only one imputation method
which allows incorporation of sensitivity parameters with multiple imputation:
the Tilted Last Occasion assumption. See the monograph for more details. This
requires providing, in addition to the marginal we want to impute, a sensitivity
parameter *ξ* to be used in exponential tilting for each imputed dataset. For
example, if $\xi \sim \text{Uniform}(0,1)$ we might do the following:

```{r TLOMI, cache = TRUE}
## Use the tilted last occasion model for imputation
imputes_tlo <- ParafacMI(Y = synthetic_data$Y, 
                         R = synthetic_data$R, 
                         chain = fit_mnar, 
                         num_impute = 20, 
                         method = "TLO", 
                         j_0 = 6, 
                         xi = runif(20))
```

# Assessing Goodness of Fit

As a sanity check, it can be useful to verify that the results of the
nonparametric model agree with a nonparametric analysis of the observed data
distribution; this ensures that the model is sufficiently flexible to model the
observed data reasonably well. The `ParafacObserved` function does this for the
marginal means of both the missing data indicators and the *observed* outcomes.
The orange line and error bars correspond to the empirical means with an
associated 95% interval, while the solid points/dark error bars correspond to
the output of the model.

```{r}
marginal_means <- ParafacObserved(fit_mnar,
  Y = synthetic_data$Y,
  R = synthetic_data$R,
  do_plot = TRUE
)
```

We see that the model is essentially in agreement with the empirical means.
