---
title: "Semi Competing Risks"
author: "Antonio R. Linero"
date: "2023-09-13"
output: html_document
editor_options: 
  markdown: 
    wrap: 80
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Setup

We illustrate the semi competing risks model by replicating the illustrations
given at <https://github.com/YanxunXu/BaySemiCompeting> by Yanxun Xu. We first
load simulated data generated at that site:

```{r, results='hide'}
library(tidyverse)
library(CausalBNPBook)
sim_data <- load("data.Rdata")

progression_time   <- full_data[,1]
survival_time      <- full_data[,2]
status_progression <- full_data[,3]
status_survival    <- full_data[,4]
trt                <- Z[,1]
X                  <- Z[,-1]
```

The package defines a wrapper function `BaySemiComp` around the code from the
github repository. It takes as arguments:

-   The progression event time (i.e., the minimum of progression time, censoring
    time, and survial time).
-   The survival event time (i.e., the minimum of the survival time and the
    censoring time).
-   The censoring indicator of progression (equal to 1 if progression is
    observed, 0 if death or censoring is observed before progression).
-   The censoring indicator of survival (indicator that death occurs before
    censoring).
-   A vector of treatment indicators.
-   A matrix of covariates to adjust for.
-   The number of warmup iterations, saved iterations, and the thinning interval
    of the MCMC scheme.
-   A grid of values of $\rho$ (the sensitivity parameter) to consider, which
    defaults to $(0.2, 0.5, 0.8)$.
-   A grid of time points to evaluate the marginal survival curve estimates at
    (defaults to an evenly-spaced grid of size $50$ covering the observed data).

The function first runs the MCMC, then computes (i) the marginal survival curves
for the treatment and control groups and (ii) an estimate of $\tau(u)$ as a
function of $u$ (the log-time). For illustrative purposes, we run the chains for
only $200$ iterations; in practice it should be run longer.

```{r fitmod, cache = TRUE}
fitted_bsc <-
  BaySemiComp(
    progression_time,
    survival_time,
    status_progression,
    status_survival,
    trt,
    X,
    100,
    100,
    1
  )
```

The following code plots the usual survival curve estimates against the output
of the BNP method with a 95% credible band:

```{r}
library(survival)
tim <- fitted_bsc$time_grid
mini.surv <- survfit(Surv(full_data[,2], full_data[,4])~ Z[,1], conf.type="none")
plot(mini.surv, col=c(1, 2), xlab="log(Time)", ylab="Survival Probability")
lines(tim, fitted_bsc$marginal_survival$fmean0_ave, col=1,lty=2)
lines(tim, fitted_bsc$marginal_survival$fquantile0[1,], col=1, lty=3)
lines(tim, fitted_bsc$marginal_survival$fquantile0[2,], col=1, lty=3)
lines(tim, fitted_bsc$marginal_survival$fmean1_ave, col=2,lty=2)
lines(tim, fitted_bsc$marginal_survival$fquantile1[1,], col=2, lty=3)
lines(tim, fitted_bsc$marginal_survival$fquantile1[2,], col=2, lty=3)
legend("topright", c("Treatment", "Control"), col = c(2,1),lty=1)
```

The following code recreates the estimate of $\tau(u)$ along a specified range
of $u$'s:

```{r}
u_range = seq(0,6,0.5)
ratio = matrix(0, length(u_range), 2)
hu_result <- fitted_bsc$causal_estimand_grid[[1]]
hu_result1 <- fitted_bsc$causal_estimand_grid[[3]]

ratio_interval=array(0, c(2, length(u_range), 2))
tmp = ifelse(hu_result[,,1]<=0.05 & hu_result[,,2]<=0.05, 1, hu_result[,,2]/hu_result[,,1])
tmp1 = ifelse(hu_result1[,,1]<=0.05 & hu_result1[,,2]<=0.05, 1, hu_result1[,,2]/hu_result1[,,1])

  for (u_index in 1:length(u_range))
  {
    ratio[u_index,1] = mean(tmp[u_index,])
    ratio_interval[, u_index, 1] = quantile(tmp[u_index,], c(0.025, 0.975))
    ratio[u_index,2] = mean(tmp1[u_index,])
    ratio_interval[, u_index, 2] = quantile(tmp1[u_index,], c(0.025, 0.975))
  }

par(mar=c(5.1,5.1,4.1,2.1))
plot(u_range, predict(loess(ratio[,1]~u_range)),"l",xlab="log(Time) (u)", ylab=expression(paste("Estimate of", " ", tau, "(u)")),ylim=c(0, 2),cex.axis=2, cex.lab=2)
lines(u_range, predict(loess(ratio_interval[1,,1]~u_range)),col=1,lty=2)
lines(u_range, predict(loess(ratio_interval[2,,1]~u_range)),col=1,lty=2)
lines(u_range, predict(loess(ratio[,2]~u_range)),col=2)
lines(u_range, predict(loess(ratio_interval[1,,2]~u_range)),col=2,lty=2)
lines(u_range, predict(loess(ratio_interval[2,,2]~u_range)),col=2,lty=2)
legend("topright", c(expression(paste(rho, "=0.2")),expression(paste(rho, "=0.8"))), col = c(1, 2),lty=c(1,1,1))
```
